Inferring the effective start dates of non-pharmaceutical interventions during \covid outbreaks

### 2020-07-12 ms/ms.tex
added Re figure

%%% Assessment
\paragraph*{Consequences of late and early effect of NPIs on real-time assessment.}

The success of non-pharmaceutical interventions is assessed by health officials using various metrics, such as the decline in the growth rate of daily cases. These assessments are made a specific number of days after the intervention began, to accommodate for the expected generation interval (i.e. average time between infection events in a chain of transmission), which is estimated at about 4-7 days~\citep{Bar-On2020}.

However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a generation interval of 4-7 days and neglect the late or early population response to the NPI.
This is illustrated in \autoref{fig:prediction} using data and parameters from Italy: a lockdown was officially ordered on Mar~10 ($\tau^*$), but its late effect on the infection dynamics starts on Mar~17 ($\hat{\tau}$).
If health officials assume that the dynamics change exactly at $\tau^*$, they would expect the number of cases to be within the red lines (posterior predictions assuming $\tau=\tau^*$).
This would lead to a significant underestimation, which might be interpreted as ineffectiveness of the NPI, leading to further escalations.
However, the number of cases would actually follow the blue lines (posterior predictions using $\tau=\hat{\tau}$), which corresponds well to the real data (stars).

%%% Fig - Prediction
\begin{figure}[h]
    \centering
	\includegraphics[width=0.4\textwidth]{../figures/Fig-prediction.pdf}
    \caption{
    \textbf{Late effective start of NPIs leads to underestimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data; line: time moving average). 
    Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution). 
    Shaded box illustrates a generation interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model underestimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are best estimates for Italy (\autoref{table:estimated-params}).
    } 
    \label{fig:prediction}
\end{figure}

@@

As different countries experiment with various intervention strategies, we expect similar shifts to occur: in some cases the population will be late to comply with the guidelines, whereas in other cases the population will adopt restrictions or relaxations even before they are formally announced.

Attempts to asses the effect of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic generation interval of \covid\citep{Bar-On2020}.
However, late and early effects such as we have inferred may bias these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:prediction}).

@@

>>>>

%%% Assessment
\paragraph*{Consequences of late and early effective start on assessment of NPIs impact.}

The reproductive number is XXX
We can infer the reproductive numbers before ($R_1$) and after ($R_2$) the implementation of NPIs from the estimated model parameters \citep{Li2020},
\begin{equation} \label{eq:Re}
\begin{aligned}
R_1 &= \alpha_1 \beta D + (1-\alpha_1) \mu \beta D, \\
R_2 &= \alpha_2 \lambda \beta D + (1-\alpha_2) \mu \lambda \beta D.
\end{aligned}
\end{equation}

The impact of NPIs can then be measured by the relative reduction in the reproductive number \citep{Flaxman2020}, $\frac{R_1 - R_2}{R1}$. 
We compared the relative reduction in the reproductive number estimated with either the fixed $\tau$ model or the free $\tau$ model.
That is, the y-axis shows the estimate assuming that NPIs started at the official date $\tau^*$, whereas the x-axis shows the estimate when inferring the effective start of the NPIs from the data. \autoref{fig:Re} demonstrates that estimates from the fixed $\tau$ model are consistently lower than estimated from a free $\tau$ model, except for Sweden and Denmark, in which the inference uncertainty in is high anyway.  
This suggests that if health officials and researchers assume that the impact of NPIs starts at the official date, then they are likely to under-estimate their impact. 
This might be interpreted as ineffectiveness of the NPI, leading to further escalations.

Health officials might also assess the impact of NPIs by comparing their expectations (e.g. model predictions) to data at a specific number of days after the intervention began. 
However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate such assessments.
This is illustrated in \autoref{fig:prediction} using data and parameters from Italy: a lockdown was officially ordered on Mar~10 ($\tau^*$), but its late effect on the infection dynamics starts on Mar~17 ($\hat{\tau}$).
If health officials assume that the dynamics change exactly at $\tau^*$, they would expect the number of cases to be within the red lines (posterior predictions assuming $\tau=\tau^*$).
This would lead to a significant underestimation, which might be interpreted as ineffectiveness of the NPI, leading to further escalations.
However, the number of cases would actually follow the blue lines (posterior predictions using $\tau=\hat{\tau}$), which corresponds well to the real data (stars).

%%% Fig - Re
\begin{figure}[h]
    \centering
	\includegraphics[width=\textwidth]{../figures/Fig_Re.pdf}
    \caption{
    \textbf{Impact of NPIs is under-estimated when assuming they start at the official date.}
    Shown are estimates of the relative reductions in reproductive numbers, which are measure of the impact NPIs have on the disease transmission.
     The y-axis shows the estimate when assuming the start of NPIs is fixed at the official date (fixed $\tau$); the x-axis show the estimate when inferring the start of NPIs from the data (free $\tau$).
    The relative reductions in $R_e$ are consistently lower for the fixed $\tau$ model.
    The reproduction numbers are $R_{1}=\alpha_1 \beta D + (1-\alpha_1) \mu \beta D$ prior to start of NPI and $R_2=\alpha_2 \lambda \beta D + (1-\alpha_2) \mu \lambda \beta D$ after start of NPI. The relative reduction is $(R{1}-R{2}) / R{1}$.
    } 
    \label{fig:Re}
\end{figure}

%%% Fig - Prediction

\begin{figure}[h]
    \centering
    \includegraphics[width=0.4\textwidth]{../figures/Fig-prediction.pdf}
    \caption{
    \textbf{Late effective start of NPIs leads to underestimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data; line: time moving average).
    Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution).
    Shaded box illustrates a generation interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model underestimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are best estimates for Italy (\autoref{table:estimated-params}).
    }
    \label{fig:prediction}
\end{figure}

@@

@@

As different countries experiment with various intervention strategies, we expect similar shifts to occur: in some cases the population will be late to comply with the guidelines, whereas in other cases the population will adopt restrictions or relaxations even before they are formally announced.
Attempts to asses the effect of NPIs~\citep{Banholzer2020,Flaxman2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic generation interval of \covid\citep{Bar-On2020}.
However, late and early effects such as we have inferred may bias these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:prediction}).
